---
title: "MLB Attendance, Home Field Advantage, and Batting Performance"
author: "Ethan Rubenstein"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Entire file takes ~2mins to run

# Read in original Retrosheet dataset and Stadium dataset, which includes attendance figures and stadium cap respectively
```{r cars}
library(tidyverse) 
# Read in 2022 Retrosheet game log which includes attendance -- https://www.retrosheet.org/gamelogs/index.html
mlb <- read.csv("gl2022.txt", header = FALSE)
mlb = mlb[!is.na(mlb$V18),]
mlb = mlb[!(mlb$V18 == 0),]
mlb = mlb[order(mlb$V18),]

# Read in stadium data with capacity from http://www.stadiumdude.com/mlb-stadium-data/ 
stadiums <- read.csv("stadiumdata.txt", sep="\t", header = TRUE)

# Remove digit group commas from stadium capacity variable and add stadium capacity to Retrosheet dataset based on stadium code
mlb$stadium_cap = as.numeric(gsub(",", "", stadiums$Capacity[match(mlb$V17, stadiums$Code)]))
mlb = mlb[!is.na(mlb$stadium_cap),]
# Calculate attendance proportion ( attendance / stadium capacity)
mlb$attend_prop = mlb$V18/mlb$stadium_cap
# Change the date format to yyyy-dd-mm to match baseballr pitch-by-pitch data for seamless merge
mlb$V1 = gsub("(\\d{4})(\\d{2})(\\d{2})$","\\1-\\2-\\3", mlb$V1)
# Create new dataframe extracting only relevant attendance and game data
clean_mlb = mlb[c("V1", "V4", "V7", "attend_prop", "V18", "stadium_cap","V17", "V2")]
# Rename column headers for ease of use
colnames(clean_mlb) = c("date", "visiting", "home", "attend_prop", "attendance", "stadium_cap", "code", "doubleheader")
```

# Scrape 2022 pitch data using  baseballr package 
```{r}
# No need to run this code, since the data has been written to file "2022data.csv"
# Get 2022 pitch-by-pitch data, data borrowed from lecture 3 R code

# library(baseballr)
# year = 2022
# temp_data = NULL
# pitch_data = data.frame(matrix(ncol = 92, nrow = 0))
# 
# # Iterate through all possible months when an MLB game could be played
# for(month in 3:3) {
#   # Determine number of days in month
#   if(month == 4 || month == 6 || month == 9 || month == 11) {
#       max_day = 30
#   } else {
#       max_day = 31
#      }
#   
#   # Iterate through all days in the current month
#   for(day in 1:max_day) {
#     if(month < 10) {
#         m = paste("0", toString(month), sep = "")
#     } else {
#         m = toString(month)
#     }
#     
#     if(day < 10) {
#         d = paste("0",toString(day),sep="")
#     } else {
#         d = toString(day)
#     }
#     
#     input_date = paste(toString(year), paste("-", paste(m, paste("-", d, sep=""), sep=""), sep=""), sep="")
#     print(input_date)
#     temp_data = try(scrape_statcast_savant(start_date = input_date, end_date = input_date))
#     if(nrow(temp_data) > 0) {
#         pitch_data = rbind(pitch_data, temp_data)
#     }
#   }
# }


# Write 2022 pitch-by-pitch to disk to avoid rerunning time-consuming code above
# pitch_data
# write.csv(pitch_data, "2022data.csv", row.names=TRUE)
```

# Read in 2022 pitch-by-pith data after
```{r}
# Read in 2022 pitch-by-pitch data
data_2022 <- read.csv("2022data.csv", header = TRUE)
```

# Calculate RE24 for each at-bat
```{r}
# Get RE24 
# Code taken from lecture 3 R code
require(devtools)
install_github("BillPetti/baseballr")
require(baseballr)
at_bat_data = run_expectancy_code(data_2022, level = "plate appearance")

# Define events to be retained in calculation of wOBA
plate_appearance_events = c("batter_interference", "double_play", "field_out", "fielders_choice", "fielders_choice_out", "force_out", "grounded_into_double_play", "sac_fly", "sac_fly_double_play", "triple_play", "field_error", "strikeout", "strikeout_double_play", "interf_def", "single", "double", "triple", "home_run", "walk", "hit_by_pitch")

# Only select regular season at-bats
at_bat_data = at_bat_data[which(at_bat_data$game_type == "R"), ]

# Identify rows to be retained if the event for the at-bat is an event defined above
indices = which(at_bat_data$events %in% plate_appearance_events)
at_bat_data = at_bat_data[indices, ]

at_bat_data = at_bat_data[order(at_bat_data$game_date),]

```
# Create attendance groups
```{r}
# Group games by attendance proportion 
# Games above Q3 in attendance proportion --> 'high' attendance group
# Games between median and Q3 in attendance proportion --> 'med-high' attendance group
# Games between Q1 and median of attendance proportion --> 'med-low' attendance group
# Games below Q1 of attendance proportion --> 'low' attendance group

# Get Q1, median, Q3 for grouping
attend_prop_sumstats = summary(clean_mlb$attend_prop)

# Create groups
clean_mlb$attend_group <- with(clean_mlb, ifelse(attend_prop >= attend_prop_sumstats[5], 'high',
                         ifelse(attend_prop >= attend_prop_sumstats[3], 'med-high', ifelse(attend_prop >= attend_prop_sumstats[2], 'med-low', 'low'))))

clean_mlb[order(clean_mlb$date),]

```
# Remove doubleheaders
```{r}
# Investigate and remove doubleheaders

# Find all doubleheader games from 2022
doubleheaders = mlb[mlb$V2 %in% c(1, 2),]
# Game 1 doubleheaders
doubleheaders = mlb[mlb$V2 %in% c(1),]
# Game 2 doubleheaders
doubleheaders = mlb[mlb$V2 %in% c(2),]
# Order by date to see that some doubleheaders are missing game 1 
doubleheaders = doubleheaders[order(doubleheaders$V1),]
doubleheaders[c("V1", "V2", "V4", "V7", "attend_prop")]

# CHA - CLE 4/20/2022 missing game 1 of doubleheader
mlb[mlb$V1 == "2022-04-20" & mlb$V4 == "CHA" & mlb$V7 == "CLE",]
# Original date for CHA - CLE 4/20/2022 game 1 was 4/19/2022 - does not exist in the dataset because it was recorded as having an attendance of 0
mlb[mlb$V4 == "CHA" & mlb$V7 == "CLE",]
# Similarly, Braves - Mets on 5/03/2022 is missing game 1, which was rescheduled from 4/04/2022
mlb[mlb$V1 == "2022-05-03" & mlb$V4 == "ATL" & mlb$V7 == "NYN",]
```
# Make team abbreviations consistent between Retrosheet and at-bat datasets, so that games can be matched
```{r}
# Team abbreviations between the datasets are not the same as the unique() functions show
sort(unique(clean_mlb$home), decreasing=TRUE)
sort(unique(clean_mlb$visiting), decreasing=TRUE)

sort(unique(at_bat_data$home_team), decreasing=TRUE)
sort(unique(at_bat_data$away_team), decreasing=TRUE)
# This function takes in the clean_mlb dataset and changes its abbreviations to match the pitch-by-pitch dataset
replace_abbrev = function(df, old, new) {
  for (x in 1:length(old)) {
    df$visiting[df$visiting == old[x]] = new[x]
    df$home[df$home == old[x]] = new[x]
  }
  return(df)
}
# Call the function with every abbreviation that needs to be changed, and what it needs to be changed to
clean_mlb = replace_abbrev(clean_mlb, c("NYN", "WAS", "TBA", "SLN", "SFN", "SDN", "NYA", "ANA", "LAN", "KCA", "ARI", "CHA", "CHN"), c("NYM","WSH", "TB", "STL", "SF", "SD", "NYY", "LAA", "LAD", "KC", "AZ", "CWS", "CHC"))

```

# Merge attendance info with pitch-by-pitch data
```{r}
# Combine attendance proportion and groups with pitch-by-pitch data
# Create gameid variable to act as an identifier for each individual game, so that overall game stats can be mapped to each pitch
clean_mlb$gameid = paste(clean_mlb$date, clean_mlb$home, clean_mlb$visiting)
at_bat_data$gameid = paste(at_bat_data$game_date, at_bat_data$home_team, at_bat_data$away_team)

# Add attendance proportion to pitch-by-pitch data
at_bat_data$attend_prop = clean_mlb$attend_prop[match(at_bat_data$gameid, clean_mlb$gameid)]
at_bat_data = at_bat_data[!is.na(at_bat_data$attend_prop),]
# Add attendance to pitch-by-pitch data
at_bat_data$attendance = clean_mlb$attendance[match(at_bat_data$gameid, clean_mlb$gameid)]
# Add attendance group to pitch-by-pitch data
at_bat_data$attend_group = clean_mlb$attend_group[match(at_bat_data$gameid, clean_mlb$gameid)]
# Add doubleheader (yes/no, which game if so) to pitch-by-pitch data
at_bat_data$doubleheader = clean_mlb$doubleheader[match(at_bat_data$gameid, clean_mlb$gameid)]


# Remove doubleheader games to avoid using the same attendance proportion for both games, for games that are missing attendance data for their first game
at_bat_data = at_bat_data[!at_bat_data$doubleheader %in% c(1, 2),]
# Remove 9th inning at-bats
at_bat_data = at_bat_data[!at_bat_data$inning == "9",]
#Verify that everything is correct - check a few at random and cross-reference with ESPN recorded attendance
at_bat_trimmed = at_bat_data[c("game_date", "home_team", "away_team", "gameid", "doubleheader", "attendance", "attend_prop", "attend_group")] 
sample_n(at_bat_trimmed, 5)
```

# Descriptive statistics - Histograms
```{r}
# Look at basic descriptive statistics for RE24 across home/away and attendance factors - Histograms

# Fig 3.2
p = ggplot(data = at_bat_data, aes(x = re24)) + geom_histogram(binwidth = 0.2)
p + facet_wrap(~inning_topbot, ) + labs(title="Histogram of RE24 per Home/Away Team", x="RE24", y="Count")
# Fig 3.4
p = ggplot(data = at_bat_data, aes(x = re24)) + geom_histogram(binwidth = 0.2)
p + facet_wrap(~attend_group) + labs(title="Histogram of RE24 per Attendance Level", x="RE24", y="Count")
# Fig 3.6
p = ggplot(data = at_bat_data, aes(x = re24)) + geom_histogram(binwidth = 0.2)
p + facet_wrap(attend_group~inning_topbot) + labs(title="Histogram of RE24 per Attendance Level and Home/Away Team", x="RE24", y="Count")
```
# Descriptive statistics - Boxplots
```{r}
# Look at basic descriptive statistics for RE24 across home/away and attendance factors - Boxplots
# Fig 3.1
ggplot(at_bat_data, aes(x=inning_topbot, y=re24, fill=inning_topbot)) + geom_boxplot(outlier.shape = NA) + ylim(-1.5, 1.5) + labs(title="RE24 by Home Team vs. Away Team", x="Inning", y="RE24") + theme(legend.position = "none")
# Fig 3.3
ggplot(at_bat_data, aes(x=attend_group, y=re24, fill=attend_group)) + geom_boxplot(outlier.shape = NA) + ylim(-1.5, 1.5)  + labs(title="RE24 by Attendance Levels", x="Attendance Group", y="RE24") + theme(legend.position = "none")
# Fig 3.5
ggplot(at_bat_data, aes(x=attend_group, y=re24, fill=inning_topbot)) + geom_boxplot(outlier.shape = NA) + ylim(-1.5, 1.5) + labs(title="RE24 by Home Team vs. Away Team and Attendance Levels", x="Attendance Group", y="RE24", fill="Inning")
```
# Summary stats - Fig 3.9
```{r}
# Mean per home/away team
aggregate(x = at_bat_data$re24 , by = list(at_bat_data$inning_topbot), FUN = mean) 
# Mean per attendance level
aggregate(x = at_bat_data$re24 , by = list(at_bat_data$attend_group), FUN = mean) 
# Mean per home/away team and attendance level   -- Fig 3.7
aggregate(x = at_bat_data$re24 , by = list(at_bat_data$attend_group, at_bat_data$inning_topbot), FUN = mean) 
```
# Descriptive stats for high-leverage scenario #1
```{r}
# High leverage scenario #1 : bases loaded with 2 outs
at_bat_data_high_lev = at_bat_data[at_bat_data$outs_when_up == 2 & !is.na(at_bat_data$on_1b) & !is.na(at_bat_data$on_2b) & !is.na(at_bat_data$on_3b),]

ggplot(at_bat_data_high_lev, aes(x=attend_group, y=re24, fill=inning_topbot)) + geom_boxplot() 
# Fig 3.8
p = ggplot(data = at_bat_data_high_lev, aes(x = re24)) + geom_histogram(binwidth = 0.5)
p + facet_wrap(attend_group~inning_topbot)  + labs(title="Histogram of RE24, 2 Outs and Bases Loaded", x="RE24", y="Count")

```
# Descriptive stats for high-leverage scenario #2
```{r}
# High leverage scenario #2: 8th inning with score differential 1 or 0
at_bat_data_high_lev_2 = at_bat_data[at_bat_data$inning == "8" & abs(at_bat_data$home_score - at_bat_data$away_score) <= 1,]

ggplot(at_bat_data_high_lev_2, aes(x=attend_group, y=re24, fill=inning_topbot)) + geom_boxplot() 
# Fig 3.9
p = ggplot(data = at_bat_data_high_lev_2, aes(x = re24)) + geom_histogram(binwidth = 0.2)
p + facet_wrap(attend_group~inning_topbot) + labs(title="Histogram of RE24, 9th Inning Score Differential <= 1", x="RE24", y="Count")
```
# Inferential stats
```{r}
# Run inferential stats tests
library(rstatix)
library(nortest)

# Only need these 3 variables to run the tests
at_bat_test = at_bat_data[c("inning_topbot", "re24", "attend_group")]
# Tests for normality 
shapiro.test(sample_n(at_bat_test[at_bat_test$attend_group == "high",], 4000)$re24)
shapiro.test(sample_n(at_bat_test[at_bat_test$attend_group == "low",], 4000)$re24)
ad.test(at_bat_test[at_bat_test$attend_group == "low",]$re24)
ad.test(at_bat_test[at_bat_test$attend_group == "high",]$re24)
# Test for equal variance
var.test(at_bat_test[at_bat_test$inning_topbot=='Top',]$re24, at_bat_test[at_bat_test$inning_topbot=='Bot',]$re24)

#Two-Sample Wilcoxon Test per attendance group
stat.test <- at_bat_test %>%
  group_by(attend_group) %>%
  wilcox_test(re24 ~ inning_topbot) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()
stat.test

# One-Sample Wilcoxon Test
wilcox.test(at_bat_test[at_bat_test$attend_group == "high" & at_bat_test$inning_topbot=='Top',]$re24, alt='l', mu=0)
wilcox.test(at_bat_test[at_bat_test$attend_group == "high" & at_bat_test$inning_topbot=='Bot',]$re24, alt='g', mu=0)

```
# Inferential stats - high leverage scenarios
```{r}
# Inferential statistics for high leverage situations

# High leverage scenario #1 : bases loaded with 2 outs
# Trim dataset to high attendance and necessary variables only
at_bat_data_high_lev = at_bat_data_high_lev[at_bat_data_high_lev$attend_group == "high",]
at_bat_data_high_lev = at_bat_data_high_lev[c("attend_group", "re24", "inning_topbot")]
at_bat_data_high_lev
# High leverage scenario #2: 8th inning with score differential 1 or 0
# Trim dataset to high attendance and necessary variables only
at_bat_data_high_lev_2 = at_bat_data_high_lev_2[at_bat_data_high_lev_2$attend_group == "high",]
at_bat_data_high_lev_2 = at_bat_data_high_lev_2[c("attend_group", "re24", "inning_topbot")]
at_bat_data_high_lev_2

# Tests null hypothesis that the home team has a higher RE24 than away team in respective scenario + high attendance
wilcox_test(re24 ~ inning_topbot, data=at_bat_data_high_lev, alternative="g")
wilcox_test(re24 ~ inning_topbot, data=at_bat_data_high_lev_2, alternative="g")

```
# Final paired-down dataset - includes Home/Away team, attendance group, and RE24 for each at-bat
```{r}
at_bat_test
```
